User guide to MSEtool

### 2018-07-25 inst/doc/MSEtool.Rmd
Update vignettes

---
title: "User guide to MSEtool"
subtitle: "Development, Testing and Implementation of Management Procedures for Data-Rich Fisheries (v1.1)"
author: "Tom Carruthers (<t.carruthers@oceans.ubc.ca>) 
         Adrian Hordyk  (<a.hordyk@oceans.ubc.ca>),
         Quang Huynh (<q.huynh@oceans.ubc.ca>)"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
  rmarkdown::pdf_vignette:
    toc: true
    number_sections: true    
vignette: >
  %\VignetteIndexEntry{User guide to MSEtool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

The package uses the Management Strategy Evaluation engine of [DLMtool](http://www.datalimitedtoolkit.org) to undertake closed-loop simulation testing of management procedures. The DLMtool package was developed at the University of British Columbia in collaboration with the Natural Resources Defense Council.

DLMtool development has been funded by the Gordon and Betty Moore Foundation, the Packard Foundation, U.S. National Oceanic and Atmospheric Administration, Fisheries and Oceans Canada, the Walton Foundation, Resources Legacy Fund, the Natural Resources Defense Council, the United Nations Food and Agricultural Organization and the California Department of Fish and Wildlife.

@@

## New Additions to this Version (v1.1)

* Assessment models (`Assess` functions)
    * Statistical Catch at Age (SCA and SCA2)
    * Delay Difference (DD_TMB)
    * State-Space Delay Difference (DD_SS)
    * Surplus Production (SP)
    * State-Space Surplus Production (SP_SS)

* Visualization
    * Standardized assessment reports
    * Diagnostics for assessment models

* Other features 
    * Harvest control rules (`HCR` functions)
    * `make_MP` function conveniently combines a __Assess__ and __HCR__ function to create a __MP__ function
    * Movement matrices for parameterizing OM
    * The MCMC outputs of iSCAM models are now used to characterise operating model uncertainty
    * iSCAM2OM and iSCAM2Data functions rip from iSCAM assessments
    * SS2OM and SS2Data functions updated for latest version of r4ss

@@

# Introduction

The MSEtool package is an extension of the DLMtool R package for implementing management strategy evaluation (MSE) simulations. Whereas DLMtool is primarily designed for testing data-limited management procedures, MSEtool provides an object-oriented framework for testing data-rich management procedures (MPs), which typically involve combining an assessment model with a harvest control rule to provide a fishery management recommendation. MSEtool management procedures are designed to be used with the operating model of DLMtool through the `runMSE` function. The user guide to the DLMtool package is accessible via

@@

The overall structure of a closed-loop simulation for a MSE in DLMtool can be described through the classes deployed in the package. An operating model (class __OM__ object) parameterizes the stock (biology), fleet, observation, and implementation dynamics of the MSE. To begin the MSE, the __OM__ object generates the data to be used for a management procedure and stores them in an object of class __Data__. A function of class __MP__ uses the __Data__ object and returns an object of class __Rec__ containing the management recommendations, e.g., effort and/or catch limits. From the management recommendations, the __OM__ object is updated and the process is successively repeated. The output for the closed-loop simulation is returned in an object of class __MSE__.

@@

MSEtool provides functions for converting the results of assessment models into DLMtool __OM__ and __Data__ objects. The SS2OM and SS2Data functions convert SS (Stock Synthesis) assessments, while iSCAM2OM and iSCAM2Data functions are provided for iSCAM assessments.

Summary additional features created by Tom... for movement.

@@

Assessment models in MSEtool are functions of class __Assess__. 
Currently, six assessment functions are available: __DD_TMB__ (Delay-Difference model and its state-space counterpart, __DD_SS__), __SP__ (Surplus Production model and its state-space counterpart, __SP_SS__), and __SCA__ (Statistical Catch at age model and another variant __SCA2__). These functions take a __Data__ object as an argument and return an object of class __Assessment__, for example,

```{r, eval = FALSE}
DD_result <- DD_TMB(1, sim_snapper, ...)
```
where `...` are model configuration arguments.

The __Assessment__ object contains the raw output from TMB estimation model, optimization output from `nlminb`, and parameter estimates with covariance matrix (slots obj, opt, and SD, respectively). If applicable, time series estimates of total biomass, vulnerable biomass, spawning stock biomass, fishing mortality and/or exploitation rate, as well as their corresponding MSY and virgin reference points. The __Assess__ object also reports likelihood values and predicted values for the data used in the assessment model.

@@

```{r, eval = FALSE}
summary(DD_result)
plot(DD_result, save_figure = TRUE, save_dir = getwd()) # By default, also saves figures in the working directory and HTML browser for viewing
```

@@

```{r, eval = FALSE}
retrospective(DD_result, 5) # Retrospective analysis going back 5 years from current year
profile_likelihood(DD_result, UMSY = seq(0.05, 0.15, 0.01), MSY = seq(1, 3, 0.1)) # Joint profile over grid of UMSY and MSY
```
<img src="DD_profile.png" width=400>

_Figure 3: Joint profile of UMSY and MSY from Delay-Difference assessment model (DD_TMB function). Contours indicate change in likelihood values from minimum (red point)._

@@

Harvest control rules in MSEtool are functions of class __HCR__, and are designed to provide different for a given assessment model. __HCR__ functions take an object of class __Assessment__ and return an object of class __Rec__. Currently, three __HCR__ functions are available as output controls (catch limits): HCR_MSY, HCR40_10, and HCR60_20. HCR_MSY prescribes the TAC to be the catch predicted with fishing at the estimated $F_{MSY}$ (the product of $F_{MSY}$ and estimated vulnerable biomass), while HCR40_10 and HCR60_20 are ramped control rules which reduce the $F_{MSY}$ catch limit based on spawning depletion.

```{r, echo = FALSE, fig.width = 5, fig.height = 4}
par(mar = c(5,5,3,1))
Brel <- seq(0, 1, length.out = 200)
plot(Brel, MSEtool::HCRlin(Brel, 0.1, 0.4), xlab = "Estimated SSB/SSB0", ylab = "TAC adjustment factor \n(proportion of FMSY catch)", main = "40-10 harvest control rule", type = "l", col = "blue")
abline(v = c(0.1, 0.4), col = "red", lty = 2)
```

@@

<img src="DD_MSE.png">

@@

Model configuration arguments can be passed through `...` when the MP is made. For example, the Ricker stock-recruitment relationship can be used in the Delay-Difference model instead of the default Beverton-Holt relationship with argument `SR = "Ricker"`:

```{r, eval = FALSE}
DD_MSY_Ricker <- make_MP(DD_TMB, HCR_MSY, SR = "Ricker")
```

@@

By default, assessment model results are not saved during the MSE. However, this feature can be turned on with argument `diagnostic = TRUE` in `make_MP`:

```{r, eval = FALSE}
DD_MSY <- make_MP(DD_TMB, HCR_MSY, diagnostic = TRUE)
```

For each MP, two diagnostic objects are saved and reported in the `DLMtool::DLMenv` environment: the system time needed to run each assessment call and the __Assessment__ objects from each assessment call. The timing information may be useful to estimate total run time for additional MSE simulations or identify iterations in which the model has trouble with convergence. The __Assessment__ objects are used to obtain convergence information, as well as retrospective estimates of biomass and mortality over successive applications.

Currently, this feature is memory-intensive and may cause performance issues or crashes. If the MSE is crashing with large number of simulations, a solution may be to run `DLMtool::runMSE()` in several batches with different seed numbers. MSE performance over all batches can be calculated by stitching the class __MSE__ objects together with `DLMtool::joinMSE()`.

The `diagnosic_AM` function plots the timing and convergence information, and `retrospective_AM` plots the assessment model's retrospective patterns (for a single simulation in the MSE).

```{r, eval = FALSE}
DD_MSY <- make_MP(DD_TMB, HCR_MSY, diagnostic = TRUE)
myMSE <- DLMtool::runMSE(OM = DLMtool::testOM, MPs = c("FMSYref", "DD_MSY", "DD_4010"))
diagnostic_AM(myMSE)
retrospective_AM(myMSE, sim = 1, MP = "DD_MSY")
```
<img src="DD_diag_AM.png">

_Figure 6. Convergence and runtime diagnostics for DD_MSY MP during the MSE. In MSEtool, convergence is defined by a positive-definite Hessian matrix. In the case of non-convergence, the previous management recommendation is used._

<img src="DD_retro_AM.png">

@@

@@

Like all other output MPs, `DLMtool::TAC()` can be used to obtain a catch limit:
```{r, eval = FALSE}
DD_MSY <- make_MP(DD_TMB, HCR_MSY)
snapper_TAC <- TAC(sim_snapper, MPs = c("DD_MSY", "DD_4010"))
plot(snapper_TAC)
```
<img src="DD_TAC.png">

_Figure 8. Stochastic TAC recommendations from the Delay-Difference model with two harvest control rules. DD_4010 prescribes zero catch because the estimated depletion is less than 0.1._

>>>>

---
title: "User guide to MSEtool"
subtitle: "Development, Testing and Implementation of Management Procedures for Data-Rich Fisheries"
author: "Tom Carruthers (<t.carruthers@oceans.ubc.ca>) 
         Adrian Hordyk  (<a.hordyk@oceans.ubc.ca>),
         Quang Huynh (<q.huynh@oceans.ubc.ca>)"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true 
vignette: >
  %\VignetteIndexEntry{User guide to MSEtool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

The [DLMtool](http://www.datalimitedtoolkit.org) package was developed at the University of British Columbia in collaboration with the Natural Resources Defense Council. DLMtool development has been funded by the Gordon and Betty Moore Foundation, the Packard Foundation, U.S. National Oceanic and Atmospheric Administration, Fisheries and Oceans Canada, the Walton Foundation, Resources Legacy Fund, the Natural Resources Defense Council, the United Nations Food and Agricultural Organization and the California Department of Fish and Wildlife.

@@

## New additions to this version (1.0.0)

* Assessment models (__Assess__ class functions)
    * Statistical Catch at Age (`SCA` and `SCA2`)
    * Delay Difference (`DD_TMB`)
    * State-Space Delay Difference (`DD_SS`)
    * Surplus Production (`SP`)
    * State-Space Surplus Production (`SP_SS`)

* Visualization functions
    * Standardized assessment reports (`plot` and `summary`)
    * Diagnostics for assessment models (`profile_likelihood` and `retrospective`)
    * Diagnostics for assessment models within closed-loop simulation (AM, assessment in MSE) (`prelim_AM`, `diagnostic_AM`, and `retrospective_AM`)

* Other features 
    * Harvest control rules (__HCR__ class functions: `HCR_MSY`, `HCR40_10`, and `HCR60_20`)
    * `make_MP` function conveniently combines a __Assess__ and __HCR__ function to create a __MP__ function
    * `simmov` function creates movement matrices for parameterizing multiple area operating models (age-independent)
    * `iSCAM2OM` and `iSCAM2Data` functions import from iSCAM assessments. The MCMC outputs of iSCAM models are now used to characterise operating model uncertainty
    * `SS2OM` and `SS2Data` functions import Stock Synthesis assessments using the r4ss package version 1.24.0 (latest CRAN version)

@@

* Age-varying movement matrices for `simmov` to condition operating models

# Introduction 
The MSEtool package is an extension of the DLMtool R package for implementing closed-loop simulation testing of management procedures, a process often referred to as management strategy evaluation (MSE). Whereas DLMtool is primarily designed for testing data-limited management procedures, MSEtool provides an object-oriented framework for creating data-rich management procedures (MPs), which typically involve combining an assessment model with a harvest control rule to provide a fishery management recommendation. MSEtool management procedures are designed to be used with the operating model of DLMtool through the `runMSE` function. The user guide to the DLMtool package is accessible via

@@

The overall structure of a closed-loop simulation for a MSE in DLMtool can be described through the classes deployed in the package. An operating model (class __OM__ object) parameterizes the stock (biology), fleet, observation, and implementation dynamics of the MSE. To begin the MSE, the __OM__ object generates the data to be used for a management procedure and stores them in an object of class __Data__. A function of class __MP__ uses the __Data__ object and returns an object of class __Rec__ containing the management recommendations, e.g., some combination of spatial closures and size, effort and catch limits. From the management recommendations, the __OM__ object is updated and the process is successively repeated as the operating model progresses over time. The output for the closed-loop simulation is returned in an object of class __MSE__, from which summary statistics can be calculated.

@@

### Additional features
MSEtool provides functions for converting the results of assessment models into DLMtool __OM__ and __Data__ objects. The `SS2OM` and `SS2Data` functions convert SS (Stock Synthesis) assessments, while `iSCAM2OM` and `iSCAM2Data` functions are provided for iSCAM assessments.

To support multiple-area operating models, the `simmov` function estimates the movement matrix based on the prescribed number of areas, biomass distribution among areas in unfished conditions, and the probability of staying in each area in the next time step.

@@

Assessment models in MSEtool are functions of class __Assess__. 
Currently, six assessment functions are available: `DD_TMB` (Delay-Difference model and its state-space counterpart, `DD_SS`), `SP` (Surplus Production model and its state-space counterpart, `SP_SS`), and `SCA` (Statistical Catch at age model and another variant `SCA2`). These functions take a __Data__ object as an argument and return an object of class __Assessment__, for example,

```{r, eval = FALSE}
DD_result <- DD_TMB(Data = Red_snapper, ...)
```
where `...` are model configuration arguments.

The __Assessment__ object contains the raw output from TMB estimation model, optimization output from `nlminb`, and parameter estimates with covariance matrix. If applicable, time series estimates of total biomass, vulnerable biomass, spawning stock biomass, fishing mortality and/or exploitation rate, as well as their corresponding MSY and virgin reference points. The __Assessment__ object also reports likelihood values and predicted values for the data used in the assessment model. The object documentation for the __Assessment__ object can be obtained through:

```{r, eval = FALSE}
class?Assessment
```

@@

```{r, eval = FALSE}
summary(DD_result)
plot(DD_result, save_figure = TRUE, save_dir = getwd()) # By default, also saves figures in the working directory and HTML browser for future viewing
```

@@

```{r, eval = FALSE}
retrospective(DD_result, nyr = 5) # Retrospective analysis going back 5 years from current year
profile_likelihood(DD_result, R0 = seq(0.75, 1.25, 0.025), h = seq(0.95, 1, 2.5e-3)) # Joint profile over grid of R0 and steepness
```
<img src="DD_profile.png" width=400>

_Figure 3: Joint profile of R0 and steepness from Delay-Difference assessment model (DD_TMB function). Contours indicate change in likelihood values from minimum (red point)._

@@

Harvest control rules in MSEtool are functions of class __HCR__, and are designed to provide different for a given assessment model. __HCR__ functions take an object of class __Assessment__ and return an object of class __Rec__. Currently, three __HCR__ functions are available as output controls (catch limits): HCR_MSY, HCR40_10, and HCR60_20. HCR_MSY prescribes the TAC to be the catch predicted with fishing at the estimated $F_{MSY}$, i.e., the product of $F_{MSY}$ and estimated vulnerable biomass, while HCR40_10 and HCR60_20 are ramped control rules which reduce the $F_{MSY}$ catch limit based on spawning depletion $SSB/SSB_0$.

```{r, echo = FALSE, fig.width = 5, fig.height = 4}
par(mar = c(5,5,3,1))
Brel <- seq(0, 1, length.out = 200)
plot(Brel, MSEtool::HCRlin(Brel, 0.1, 0.4), xlab = expression(Estimated~~SSB/SSB[0]), ylab = "TAC adjustment factor \n(proportion of FMSY catch)", main = "40-10 harvest control rule", type = "l", col = "blue")
abline(v = c(0.1, 0.4), col = "red", lty = 2)
```

@@

<img src="DD_Tplot2.png">

@@

Model configuration arguments can be passed through `...` in `make_MP` when the MP is made. For example, the Ricker stock-recruitment relationship can be used in the Delay-Difference model instead of the default Beverton-Holt relationship and steepness can be fixed instead of esimated by passing argument `SR = "Ricker"` and `fix_h = TRUE`, respectively:

```{r, eval = FALSE}
DD_MSY_Ricker_fixsteep <- make_MP(DD_TMB, HCR_MSY, SR = "Ricker", fix_h = TRUE)
```

@@

Three functions are designed to evaluate the performance of __Assess__ models in MSE: `prelim_AM`, `diagnostic_AM`, and `retrospective_AM` (where _AM_ stands for "assessment in MSE").

`prelim_AM` is designed to evaluate the configuration of the assessment model before running the closed-loop simulation. Given an operating model, this function generates data from the historical period of the MSE, applies the assessment model to those data, then returns the convergence rate of the model. A list of __Assessment__ objects are returned and can be used for troubleshooting. Additional arguments are passed to the __Assess__ function via `...`:

```{r, eval = FALSE}
prelim_AM(DLMtool::testOM, DD_TMB, ...)
#> Generating Hist object from OM object via runMSE:
#> Loading operating model
#> Optimizing for user-specified movement
#> Optimizing for user-specified depletion
#> Calculating historical stock and fishing dynamics
#> Calculating MSY reference points
#> Calculating B-low reference points
#> Calculating reference yield - best fixed F strategy
#> Returning historical simulations
#> Running DD_TMB with 48 simulations for DLMtool::testOM.
#> Assessments complete.
#> Total time to run 48 assessments: 1.2 seconds
#> 2 of 48 simulations (4.2%) did not converge.
#> See simulation number: 11 21
```

By default, assessment model results are not saved during the MSE (called by `DLMtool::runMSE`). However, this feature can be turned on with argument `diagnostic` in `make_MP`:

```{r, eval = FALSE}
DD_MSY <- make_MP(DD_TMB, HCR_MSY, diagnostic = "full")
```

For each MSEtool MP, two diagnostic objects can be saved and reported in the `DLMtool::DLMenv` environment: convergence diagnostics and the __Assessment__ objects from each assessment call. The first can be obtained with `diagnostic = "min"`, while both are returned with `diagnostic = "full"`.

MSEtool MPs may be memory-intensive and cause performance issues or crashes. With large number of simulations, a solution may be to run `DLMtool::runMSE()` in several batches with different seed numbers. MSE performance over all batches can be calculated by stitching the class __MSE__ objects together with `DLMtool::joinMSE()`.

The `diagnosic_AM` function plots the convergence information, and `retrospective_AM` plots the assessment model's retrospective patterns (for a single simulation in the MSE).

```{r, eval = FALSE}
DD_MSY <- make_MP(DD_TMB, HCR_MSY, diagnostic = "full")
DD_4010 <- make_MP(DD_TMB, HCR40_10)
myMSE <- DLMtool::runMSE(OM = DLMtool::testOM, MPs = c("FMSYref", "DD_MSY", "DD_4010"))
diagnostic_AM(myMSE)
retrospective_AM(myMSE, sim = 1, MP = "DD_MSY")
```
<img src="DD_diagnostic_AM.png">

_Figure 6. Convergence diagnostics (return by TMB objects and nlminb function) for DD_MSY MP during the MSE. Here, the assessment model was applied beginning in year 50 of the MSE and repeated every 4th year. In MSEtool, convergence is defined by a positive-definite Hessian matrix (top middle figure). In the case of non-convergence, the previous management recommendation is used._

<img src="DD_retrospective_AM.png">

@@

Once it is determined that the MSEtool assessment models are performing adequately (no major convergence issues), then the performance of the MPs that use these models can be evaluated with the DLMtool package.

@@

Like all other output MPs, `DLMtool::TAC()` can be used to obtain a catch limit:
```{r, eval = FALSE}
DD_MSY <- make_MP(DD_TMB, HCR_MSY)
DD_4010 <- make_MP(DD_TMB, HCR40_10)
snapper_TAC <- TAC(Red_snapper, MPs = c("DD_MSY", "DD_4010"))
plot(snapper_TAC)
```
<img src="DD_TAC.png">

_Figure 8. Stochastic TAC recommendations from the Delay-Difference model with two harvest control rules. When the 40-10 harvest control rule is applied, the prescribed TAC is close to zero because the estimated terminal year biomass is 0.13 in the assessment model._
